---
title: "Final Project - Draft Analysis"
author:  Blanka Balazs, Simon Fernezelyi, Thea Goslicki
data: "Due 8 January 2022, 23h59"
output:
  bookdown::pdf_document2: default
urlcolor: blue
bibliography: references.bib
csl: american-political-science-association.csl
editor_options: 
  markdown: 
    wrap: 72
---
##Potential research question: 
To what extend had the implementation of Fox News an effect on the US voting behavior during the presidential elections 1996 & 2000?


##Short literature review 
Given existing literature, explain in a few sentences what is your contribution with this project. (What has been done in the literature, what is missing and how you make a contribution).

-->idea: http://exedra.esec.pt/docs/N4/10C_Rui-Antunes_pp_145-170.pdf


##hypotheses
Write initial theoretical expectations and hypotheses - so that you know what kind of analyses you would like to run.

This paper will follow Lazarsfeld et al.'s (1944) approach and main Hypotheses,called: 

HP1: The act of voting is an individual act, affected mainly by the personality of the voter and his exposure to the media.

In the case of our research and our available data, we categorize the variables about socioeconomic background of the people as personality trades and the exposure to media via the existence or non-existence of Fox News. 

Sociological model of voting Kriesi's (2008)

HP2: People’s voting behavior is just shaped by long-term factors (socialization: controlvaribales). Independent from fox news.

Rational expectation theories (Bray and Kreps 1987)

HP3: Voters often filter out reporting bias without being influenced. The implementation of Fox News will not effect election results.

Behavioral Theories (De Marzo, Vayanos, and Zwiebel 2003)

HP4:assert that media influence on voters exists.



Load the data you have selected
Run any preprocessing that you need to conduct to prepare your data for analysis (e.g. renaming, recoding of variables, creating new variables, merging files, tackling NAs etc.) 
Produces summary statistics (e.g. tables and plots) of the relevant dependent and independent variables
Conduct the main regression of interest for the project. Discuss initial findings and problems.



\newpage

##Load the data you have selected
```{r}
data <- read.csv("foxnews.csv")
data$state <- factor(data$state)
```


Mean gop by states

```{r}

gop2000_by_states <- tapply(data$gopvoteshare2000, data$state, mean, na.rm = TRUE)
gop2000_by_states
```

*we could make this into a nice table*

Other years

```{r}
gop1996_by_states <- tapply(data$gopvoteshare1996, data$state, mean, na.rm = TRUE)
gop1992_by_states <- tapply(data$gopvoteshare1992, data$state, mean, na.rm = TRUE)

# this dataframe contains the mean gop for the three years (1992, 1996, 2000) by state
gop_by_states <- data.frame(gop1992_by_states, gop1996_by_states, gop2000_by_states)

#sort it for easier use
gop_by_states <- gop_by_states[order(gop_by_states$gop2000_by_states), ]

#plot the results
plot(1, type = "n", xlab = "",
     ylab = "", xlim = c(1992, 2003), 
     ylim = c(0.3, 0.8))
for(x in 1:nrow(gop_by_states)){
  lines(as.numeric(gop_by_states[x, ]) ~ c(1992, 1996, 2000))
  text(2000.5+((x-1)%%6)/2, gop_by_states$gop2000_by_states[x], row.names(gop_by_states)[x])
}

lines(c(mean(gop1992_by_states, na.rm=T), mean(gop1996_by_states), mean(gop2000_by_states)) ~ c(1992, 1996, 2000), 
      lty=2, 
      col="red",
      lwd=3)
```

```{r}
cor(data[,c(3, 4, 5, 6, 7, 12)])
```
```{r}

gop1992_df<- data[ , c("state","town","college1990", "male1990", "black1990", "hisp1990", "income1990","subrf2000", "gopvoteshare1992")] 
gop1996_df<- data[ , c("state","town","college1990", "male1990", "black1990", "hisp1990", "income1990","subrf2000", "gopvoteshare1996")] 
gop2000_df<- data[ , c("state","town","college1990", "male1990", "black1990", "hisp1990", "income1990","subrf2000", "gopvoteshare2000")] 

plot(gop2000_df$gopvoteshare2000, gop2000_df$subrf2000, 
     pch = 16, col = "black",
      xlab = "Two-party vote share for the Republicans (2000 Presidential election)",
      ylab = "Share of Fox News subscribers by town")


tapply(gop2000_df$gopvoteshare2000, gop2000_df$subrf2000, mean,na.rm=T)
lines()
```
```{r}
gop_by_s92<- data.frame(gop1992_by_states)
gop_by_s96<- data.frame(gop1996_by_states)
gop_by_s20<- data.frame(gop2000_by_states)

plot(gop2000_df$state, gop2000_df$subrf2000,
     pch = 16, col="black",
      xlab = "State",
      ylab = "Share of Fox News subscribers")

```
```{r}
towns_w_foxnews<- subset(data, (subset = subrf2000 >0))
towns_no_foxnews<-subset(data, (subset = subrf2000 ==0))

means.all.elections.wFN<-colMeans(towns_w_foxnews[c('gopvoteshare1992','gopvoteshare1996','gopvoteshare2000')], na.rm = T)
means.all.elections.wFN

means.all.elections.noFN<-colMeans(towns_no_foxnews[c('gopvoteshare1992','gopvoteshare1996','gopvoteshare2000')], na.rm = T)
means.all.elections.noFN

years<- c(1992, 1996, 2000)

plot(years, means.all.elections.noFN, pch = 16, col = "black",
xlim = c(1992, 2000), ylim = c(0.4, 0.55), xlab = "Election years",
ylab = "Average vote share for the Republicans ",
main = "Difference in Average vote share across towns with/without fox news")

points(years, means.all.elections.wFN,pch = 16, col = "red")


lines(years, means.all.elections.noFN, type = "c")
lines(years, means.all.elections.wFN, type = "c", col="red")
text(1994,0.51, "Towns with Fox News", col="red")
text(1994,0.46, "Towns without Fox News")

```

##A Question I (Thea) would like to do



Our hypotheses stating, that it might be possible that education, gender, and people with different race reacting differently to the implementation of Fox News. 
It could be also possible that voters with different incomes get effected differently by the input of Fox News.

We will evaluate whether these hypotheses are supported by finding the differences in sample average treatment effects by education, gender, race and income in towns with access to Fox News and towns without access to Fox News. 

Furthermore,Computeing the average change of votes for republican (during presidential elections) among towns before and after the availability of Fox News could be an crucial indicator to understand the effect.




\newpage

# References
