---
title: "Revising the Fox News Effect"
author:  Blanka Balazs, Simon Fernezelyi
data: "May 06, 2023"
header-includes:
 \usepackage{float}
 \floatplacement{figure}{H}
fig_caption: yes
output:
  bookdown::pdf_document2: default
urlcolor: blue
abstract: "The conservative news channel Fox News was founded in 1996 and gradually implemented in various US cities. The 2000 Presidential election was the first time in history where a significant proportion of the US population already had access to this TV channel for several years while other cities having no access. This study observes the vote shares for the Republicans of the Presidential election in years 1992, 1996 and 2000 depending on socio-economic factors and the share of Fox News subscribers in different areas. We ask the question, to which extent did the introduction of Fox News affected US voting behavior in the presidential elections between 1996 and 2000. The analysis yields revealing results, but the idea of the highly influential TV channel in terms of political outcomes, the so-called 'FOX NEWS EFFECT', had to be revised, as the analysis showed no significance in this area."
bibliography: citations.bib
---

# **Introduction**

The media landscape has changed dramatically in the last few decades. Media have an increasingly important role in our modern society where access to information is not a privilege anymore. As a consequence of media's constant presence in citizen's life, the effects of media and to what extent media bias can influence people's party choices have also become a crucial question. This research paper aims to focus on the potential influence of media on people's party choices. To examine the problem the research paper will focus on the presence of Fox News in cities in the US and how it influenced people's support for the Republican party. Therefore, the research question will be the following: to what extent had the implementation of Fox News affected the US voting behavior in the presidential elections between 1996 and 2000? In the case of our research and our available data, we categorize the variables about the socioeconomic background of the people as personality trades and the exposure to media via the existence or non-existence of Fox News. Therefore, the research question will be answered through two hypotheses: H1:Fox news and the socioeconomic background effect the behavior of the electorate.

# Literature Review

In recent decades, there has been increasing attention on the role of media during political campaigns and its effects on the actual outcomes of elections. It is more widespread to include factors beyond the traditional social and economical models. However, it is still debated within academia to what extent news media influence the behavior of the electorate. A potential answer to the question is that news media does not have an actual effect on people's voting behavior. It is called the sociological model of voting behavior which states that people's voting behavior is determined by long-term factors.

This theory is supported by Lazarsfeld, Berelson, and Gaudet's [@Lazarsfeld1968] research analysis whose results show that the effect of media on electoral decisions is minimal but people belonging to different social groups is what determines their voting behavior. Kriesi, Grande, and Lachat [@Kriesi2008] also argue people's voting behavior is determined by their belonging to different cleavages. However, instead of the old cleavages, there is a new social division within society that defines people's voting behavior which is the distinction between the winners and losers of globalization [@Kriesi2008]. Losers of globalization feel that their social status is protected by the nation-state and therefore they are strongly connected and identify themselves with the national community and that is what determines their voting behavior [@Kriesi2008]. In contrast, the winners of globalization benefit from open borders, they have more opportunities and therefore they vote differently [@Kriesi2008]. However, some scholars argue that media has a relatively insignificant, but not at all inherent effect on voting behavior [@alotaibi2013media].

# Hypotheses

Therefore, based on the existing literature the two hypotheses of the analysis, as mentioned above, will be the following:

**HP1:** Fox news and the socioeconomic background affects the behavior of the electorate.

Sociological model of voting Kriesi's [@Kriesi2008]: People's voting behavior is just shaped by long-term factors (socialization).

# General Overview Over the Data

This paper uses data that was collected in 1992, 1996 and 2000. The dataset [@DellaVigna2007] contains town-based data on two-party vote share for the Republicans, socio-cultural details and share of Fox news subscribers.

```{r, echo=FALSE}
data <- read.csv("foxnews.csv")

ntowns <- length(data$town)
nstates <- length(unique(data$state))

```

The dataframe holds data from `r ntowns` towns across `r nstates` American states.

In order to have a rough overview of the vote share among states, the *state* variable is converted into a factor variable. With this modification the average two-party vote share for the Republicans at the different presidential elections (1992, 1996, 2000) in the different states could be calculated as shown in Figure \ref{fig:avgVoteShare}.

```{r avgVoteShare, warning=FALSE, fig.cap="Average two-party vote share for the Republicans in 2000 in the different states.", echo=FALSE}
data$state <- factor(data$state)

gop1992_by_states <- tapply(data$gopvoteshare1992, data$state, mean, na.rm = TRUE)
gop1996_by_states <- tapply(data$gopvoteshare1996, data$state, mean, na.rm = TRUE)
gop2000_by_states <- tapply(data$gopvoteshare2000, data$state, mean, na.rm = TRUE)

plotdata <- data.frame(
  state = names(gop2000_by_states),
  gop1992_by_states,
  gop1996_by_states,
  gop2000_by_states
)

library(ggplot2)
library(reshape2)

data_melted <- reshape2::melt(plotdata, id.vars = "state")

# Create the bar plot
ggplot(data_melted, aes(x = reorder(state, value, na.rm=T), y = value, fill = variable)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "State", y = "Republican vote share", fill = "Year") +
  scale_fill_discrete(name = "Year", labels = c("1992", "1996", "2000")) +
  theme_bw() +
  theme(text=element_text(family="serif"))

```

From these values it is already obvious that there is quite a big variation between the two-party vote share for the Republicans in different states. Highest and lowest vote share is measured in Utah (`r max(gop2000_by_states)`) and Rhode Island (`r min(gop2000_by_states)`) respectively.

The plot also shows general trends in the different states over time. On average there has been a slight decrease in republican vote share in the period of 1992 and 1996, however there has been differences in different region. For example Main experienced a huge drop while in Idaho a slight increase is visible.

On the other hand in the time period of 1996 and 2000 there has been a strong increase (on average from `r mean(gop1996_by_states)` to `r mean(gop2000_by_states)` ). An important note is that there has been no single state where the republican vote share has decreased in this time period.

## Socio-Cultural Descriptors of the Data Set

As an initial step, since most descriptive values are measured in the range of [0, 1], the *income1990* variable is normalized in the same region as well. In our hypotheses the independent variables are the different socio-cultural descriptors such as the education level, the ethnic composition and the income share. The varying dimension is the so-called "low-high dimension" which describes that 0 - 100 per cent of the population is black or Hispanic, the the whole population has low or high income or education level.

```{r, echo=FALSE}
descriptors <- summary(data[,c(3, 4, 5, 6, 7)])[c(1, 3, 4, 6),]

knitr::kable(
  descriptors,
  "simple",
  col.names = c("Collage eduacted", "Male", "Black", "Hispanic", "Average income"), 
  caption = "Basic descriptors of the independent variables",
  digits = 2
)
```

```{r independentVariables, fig.cap="Independent variables in the study.", echo=FALSE}
#normalizing income to the range of [0, 1] (was originially in [0, 15])
data[7] <- data[7]/15

par(family = "serif")
boxplot(data[,c(3, 4, 5, 6, 7)], ylab="Low-high dimension", names = c("Collage eduacted", "Male", "Black", "Hispanic", "Income"))
```

The Table 1 and Figure \ref{fig:independentVariables} shows the general quality of the independent variables of the data set. It is visible that the median of male population in the towns are around 0.5, however there are more towns where there are an extremity towards a high male population share than to women. The black and Hispanic population of the towns in the study is generally low, but there are cases with especially high rate from both ethnic groups. The median of education and the income is below 0.2 and the later one has a higher expansion.

## Change of Vote-Share Over Time

To have a more detailed overview, the differences in the two-party vote share for Republicans between the observed election is calculated. The histograms in Figure \ref{fig:change} indicate that there has been a slight decrease of Republican vote share between 1992 and 1996 which was replaced by a much higher increase between 1996 and 2000. It again shows that while in the first period the decrease was rather moderate, the increase in the later period was exclusive to all observed states.

```{r change, fig.show="hold", out.width="50%", fig.cap="Change in Republican vote share between the three observed elections.", echo=FALSE}

change1992_1996 <- gop1996_by_states - gop1992_by_states
change1996_2000 <- gop2000_by_states - gop1996_by_states

par(family = "serif")
hist(change1992_1996, 
     xlim = c(-0.2, 0.2),
     main="Distribution of change betwen 1992 and 1996",
     xlab = "Difference between elections"
     )
hist(change1996_2000, xlim = c(-0.2, 0.2),
     main="Distribution of change betwen 1996 and 2000",
     xlab = "Difference between elections")


```

## Correlation Analysis

The dataset contains various measures of the towns in question. In order to get a sense what could influence the vote share the correlation of these factors and the vote share is calculated.

```{r, echo=FALSE}
cordata <- cor(data[12], data[,c(3, 7, 5, 6)])
rownames(cordata) <- c("Republican vote share")
knitr::kable(
  cordata,
  "simple",
  col.names = c("Level of education", "Average income", "Share of black people", "Share of hispanic people"), 
  caption = "Correlation between Republican vote share and different independent variables.",
  digits = 4,
)
```

The correlation analysis of Table 2 shows that there are a few factors that have an close to zero correlation to the vote share for the Republicans, such as the income or the education level. The share of males in the population has a direct but moderate correlation with the vote share as it is a positive number and reaching \~0.2. The share of different ethnic categories (Hispanic or Black) in the population has an inverse relationship to the vote share, being both negative numbers. The highest correlation is between the black population and the vote share (-0.331) suggesting a moderate negative linear relationship between these two variables. The negative sign indicates that as the percentage of the black population increases in a given district, the vote share tends to decrease. The magnitude of the correlation coefficient indicates that this relationship is not extremely strong, but it is still statistically significant. It is however important to note that correlation does not imply causation.

## Fox News Subscribers in Different Towns

```{r scatterPlot, fig.cap="Change in vote share for republicans between the presidential elections of 1996 and 2000 versus the share of Fox News subscribers in 2000 by town", echo=FALSE}
data$diff_1996_2000 <- data$gopvoteshare2000 - data$gopvoteshare1996



par(family = "serif")

plot(data$subrf2000, data$diff_1996_2000,
     pch = 16, col = "black",cex=0.5,
      ylab = "Change of vote share for the Republicans",
      xlab = "Share of Fox News subscribers by town")
abline(h=0, col="grey", lty = 2 )
abline(h=mean(data$diff_1996_2000), col="red", lty = 1)
text(0.8, 0.15, "mean", col="red")


```

Figure \ref{fig:scatterPlot} gives an first overview between the change in vote share for republicans between the presidential election of 1996 and 2000 and the share of Fox News subscribers by town in 2000. Three fundamental things, which are a basis for further analysis, can be read from the scatterplot: (1) no clear direction of the data is discernible at first glance; (2) cities that did not have access to Fox News in 2000 range across the spectrum of Republican support, from no Fox News and low support for Republicans to no Fox News and very high vote share for Republicans. (3) it looks like (purely visual 'first analysis') the majority of cities are around the cut point (0.5/0.4), which means there are a lot of cities that are relatively split between Democrats and Republicans and between having or not having Fox News. Since most of the data is expressed in proportions, this also means that there will be many values in the middle.

```{r foxSubscribers, fig.cap="Share of Fox News subscribers by state in 2000", echo=FALSE, warning=FALSE}
subrf2000_by_states <- data.frame(state = c("Ak","Al","Ar","Ca","Ct","Hi","Ia","Id","Ma","Me","Mi","Mn","Mo","Mt","Nd","Nh","Nj","Ny","Oh","Pa","Ri","Sc","Tn","Ut","Va","Vt","Wi","Wy")
, values = tapply(data$subrf2000, data$state, mean, na.rm = TRUE))
library(usmap)
plot_usmap(data = subrf2000_by_states) + 
  scale_fill_continuous(low = "white", high = "#00aaff", name = "Share of subscribers") +
  theme(legend.position = "right") +
  theme(text=element_text(family="serif"))

```

To get a better picture of the implementation of Fox News in the USA, the share of Fox News subscribers by state in 2000 got plotted. The differences are immense. New Hampshire does not have any access to Fox News in 2000 yet, also in Vermont, Massachusetts and South Carolina is the access ratio to the Fox News Channel almost non existence. On the other hand is the channel already very accessible in states like New Jersey,Wyoming and Alaska. According to the bar plot has the biggest proportion of Fox News subscribers. However, due to the fact, that the bar plot does not represents the number of Fox News subscribers, only the the proportion of citizens with a subscription per town it does not mean that Hawii does have the highest amount in a quantitative sense.

# The Fox News Effect

```{r diffInDiffFoxnews, fig.cap="Differences in Vote Share Across Towns With/Without Fox News", echo=FALSE, warning=FALSE}

towns_w_foxnews<- subset(data, (subset = subrf2000 >= 0.75))
towns_some_foxnews<- subset(data, (subset = subrf2000 > 0  & subrf2000 < 0.75))
towns_no_foxnews<-subset(data, (subset = subrf2000 ==0))

means.all.elections.wFN<-colMeans(towns_w_foxnews[c('gopvoteshare1992','gopvoteshare1996','gopvoteshare2000')], na.rm = T)


means.all.elections.someFN<-colMeans(towns_some_foxnews[c('gopvoteshare1992','gopvoteshare1996','gopvoteshare2000')], na.rm = T)


means.all.elections.noFN<-colMeans(towns_no_foxnews[c('gopvoteshare1992','gopvoteshare1996','gopvoteshare2000')], na.rm = T)


years<- c(1992, 1996, 2000)

par(family = "serif")
plot(years, means.all.elections.noFN, pch = 16, col = "black",
  xlim = c(1992, 2000), ylim = c(0.2, 0.7), xlab = "Election years",
  ylab = "Average vote share for the Republicans ")

points(years, means.all.elections.wFN,pch = 16, col = "red")
points(years, means.all.elections.someFN,pch = 16, col = "blue")

abline(v=1996, lty= 2, col= "grey")

lines(years, means.all.elections.noFN, type = "c")
lines(years, means.all.elections.someFN, type = "c", col= "steelblue2")
lines(years, means.all.elections.wFN, type = "c", col="red")

text(1996, 0.59,"Implementation of Fox News", col="grey")
legend(x = 1992, y = 0.4, bty = "n",
       legend = c("Towns with Fox News", "Towns with some Fox News", "Towns without Fox News"),
       col = c("red", "steelblue2", "black"), pch = 19,
       pt.cex = c(0.5))


```

To Test the hypotheses of the research and to measure the effect of the treatment, it is crucial to define the independent variable of the paper: access to Fox News by the year 2000. As mentioned above Fox News got implemented in 1996. Therefore, the cities without any access to Fox News in 2000 got characterized as 'Towns without Fox News'. Cities that had a share of Fox News subscribers over 0.75 got defined as 'Towns with Fox News', due to the fact, that 75% of the inhabitants is a clear majority, especially when we consider that people live in households with other persons. The data set does not elaborate the information on whether children or other people who are not allowed to vote in the states ( like prisoners) are included or not, however a threshold of 0.75 this circumstance cushions instead a threshold of 1. The third subset will be 'Towns with some Fox News', in particular, all towns between 0 and 0.75 will be part of it.

Figure \ref{fig:diffInDiffFoxnews} illustrates that in all cities, no matter how much access to Fox News the various cities had, there was an increase in Republican voters in 2000, compared to the 1996 presidential election. Nevertheless, a staggered effect of the different groups can be seen, even if it is small.Cities with Fox News voted for the Republican president by a larger percentage in both 1996 (the year Fox News was implemented in some cities) and the 2000 presidential election. Cities without any Fox News access were the least likely to vote in favor of Republicans, even though the percentage of Republican voters nevertheless increased in 2000.

## Sample average treatment effect ('Fox News Effect') among socioeconomic variables

The hypothesis states, that it might be possible that socioeconomic backgrounds like education and race of people influence a potential 'Fox News Effect' differently. It could be also possible that voters with different incomes get affected differently by the input of Fox News. We will evaluate whether there is a difference in the effects of Fox News on the Republican vote share between cities where socioeconomic variables differ. Furthermore, the average change of votes for Republicans (during presidential elections) among towns before and after the availability of Fox News could be a crucial indicator to understand the effect, therefore is computed. To measure the differences between cities a difference in differences research design will be used. 

```{r diffInDiff_black, fig.cap="Differences in Vote Share in Towns with High Black Population", echo=FALSE, warning=FALSE}

##towns with high proportion of a black population

high_black_pop.t_w_foxnews<- subset(towns_w_foxnews, subset = black1990 >= 0.4)

high_black_pop.t_no_foxnews<- subset(towns_no_foxnews, subset = black1990 >= 0.4)

high_black_pop.t_some_foxnews<- subset(towns_some_foxnews, subset = black1990 >= 0.4)



means_high_black_pop.t_w_foxnews<-colMeans(high_black_pop.t_w_foxnews[c('gopvoteshare1992','gopvoteshare1996','gopvoteshare2000')], na.rm = T)

means_high_black_pop.t_no_foxnews<-colMeans(high_black_pop.t_no_foxnews[c('gopvoteshare1992','gopvoteshare1996','gopvoteshare2000')], na.rm = T)

means_high_black_pop.t_some_foxnews<-colMeans(high_black_pop.t_some_foxnews[c('gopvoteshare1992','gopvoteshare1996','gopvoteshare2000')], na.rm = T)


par(family = "serif")
plot(years, means_high_black_pop.t_w_foxnews, pch = 16, col = "aquamarine3",
xlim = c(1992, 2000), ylim = c(0.2, 0.7), xlab = "Years of presidential elections",
ylab = "Average vote share for Republicans")

points(years, means_high_black_pop.t_some_foxnews,pch=16, col = "darkseagreen4")
points(years, means_high_black_pop.t_no_foxnews,pch = 16, col = "mediumorchid1")


lines(years, means_high_black_pop.t_w_foxnews, type="c", col="aquamarine3")
lines(years, means_high_black_pop.t_some_foxnews,type="c", col = "darkseagreen4")
lines(years, means_high_black_pop.t_no_foxnews,type="c", col = "mediumorchid1")

legend(x = 1992, y = 0.7, bty = "n",
       legend = c("Towns with Fox News", "Towns with some Fox News", "Towns without Fox News"),
       col = c("aquamarine3", "darkseagreen4", "mediumorchid1"), pch = 19,
       pt.cex = c(0.5))

```

<<<<<<< HEAD
The first thing we looked at was whether the introduction of Fox News had a bigger/smaller impact on black people in the USA. The cities were divided into subsets, to take into account different proportions of black people in the cities. For the subdivision of cities and their black population, the different quantiles were not looked at, which is the case in the following SATE calculations. Due to the white majority population in the states, the third quantile of the proportion of black people would have been 0.013411 but the maximum would have been 0.990427. The difference between 1% and 99% would have been too serious to categorize all cities in between as having a high black population. That's why cities with a percentage above 0.40 were considered to have a high black population. As can be seen in the plot, only the line from the cities with a high black population share has a similar shape to the lines in the plot with the general trend effect. In the cities with a low proportion or none at all, the estimated treatment effect goes in a negative direction.

```{r diffInDiffIncome, fig.cap="Differences in Vote Share in High and Low Income Towns", echo=FALSE, warning=FALSE}
##SATE town with high proportion of rich people
#No Thea's 
=======
Figure \ref{fig:diffInDiff_black} focuses on the changes on the Republican party's vote share in cities where the black population tend to be high and we measured whether the introduction of Fox News had a bigger/smaller impact on black people in the USA. The African American population making up 15% of the US population and in the analysis cities with a percentage above 0.40 were considered to have a high black population. The results show that in those cities where the black population is considered to be high and people had access to Fox News the Republican vote share increased. However, in cities where people did not have access to Fox News, or the access was more limited the Republican vote share was stagnatig between 1996 and 2000. 

```{r diffInDiffincome, fig.cap="Differences in Vote Share in High and Low Income Towns", echo=FALSE, warning=FALSE}
## towns with high proportion of rich people
>>>>>>> 2cc3803087b13bf2a496dd0f7d4db66cbc1ce6e8

high_income.t_w_foxnews<- subset(towns_w_foxnews, subset = income1990 >= 3.420)
high_income.t_no_foxnews<- subset(towns_no_foxnews, subset = income1990 >= 3.420)
high_income.t_some_foxnews<- subset(towns_some_foxnews, subset = income1990 >= 3.420)

means_high_income.t_w_foxnews<-colMeans(high_income.t_w_foxnews[c('gopvoteshare1992','gopvoteshare1996','gopvoteshare2000')], na.rm = T)

means_high_income.t_no_foxnews<-colMeans(high_income.t_no_foxnews[c('gopvoteshare1992','gopvoteshare1996','gopvoteshare2000')], na.rm = T)

means_high_income.t_some_foxnews<-colMeans(high_income.t_some_foxnews[c('gopvoteshare1992','gopvoteshare1996','gopvoteshare2000')], na.rm = T)

low_income.t_w_foxnews<- subset(towns_w_foxnews, subset = income1990 <= 1.500)
low_income.t_no_foxnews<- subset(towns_no_foxnews, subset = income1990 <= 1.500)
low_income.t_some_foxnews<- subset(towns_some_foxnews, subset = income1990 <= 1.500)

means_low_income.t_w_foxnews<-colMeans(low_income.t_w_foxnews[c('gopvoteshare1992','gopvoteshare1996','gopvoteshare2000')], na.rm = T)

means_low_income.t_no_foxnews<-colMeans(low_income.t_no_foxnews[c('gopvoteshare1992','gopvoteshare1996','gopvoteshare2000')], na.rm = T)

means_low_income.t_some_foxnews<-colMeans(low_income.t_some_foxnews[c('gopvoteshare1992','gopvoteshare1996','gopvoteshare2000')], na.rm = T)


par(family = "serif")
par(mfrow = c(1, 2))
plot(years, means_high_income.t_w_foxnews, pch = 16, col = "darkgreen",
xlim = c(1992, 2000), ylim = c(0.2, 0.7), xlab = "Years of presidential election",
ylab = "Average vote share for Republicans",
main = "High-Income Towns")

points(years, means_high_income.t_no_foxnews,pch = 16, col = "orange")
points(years, means_high_income.t_some_foxnews,pch = 16, col = "brown")


lines(years, means_high_income.t_w_foxnews, type="c", col="darkgreen")
lines(years, means_high_income.t_no_foxnews,type="c", col = "orange")
lines(years, means_high_income.t_some_foxnews,type="c", col = "brown")

legend(x = 1992, y = 0.4, bty = "n",
       legend = c("Towns with Fox News", "Towns with some Fox News", "Towns without Fox News"),
       col = c("darkgreen", "orange", "brown"), pch = 19,
       pt.cex = c(0.5))

plot(years, means_low_income.t_w_foxnews, pch = 16, col = "darkgreen",
xlim = c(1992, 2000), ylim = c(0.2, 0.7), xlab = "Years of presidential elections",
ylab = "Average vote share for Republicans",
main = "Low-Income Towns")

points(years, means_low_income.t_no_foxnews,pch = 16, col = "orange")
points(years, means_low_income.t_some_foxnews,pch = 16, col = "brown")


lines(years, means_low_income.t_w_foxnews, type="c", col="darkgreen")
lines(years, means_low_income.t_no_foxnews,type="c", col = "orange")
lines(years, means_low_income.t_some_foxnews,type="c", col = "brown")




```

<<<<<<< HEAD
Looking at income and the potential sample average treatment effect we can see, that only for persons with high income (towns where the average income is over or equals 3.420\$) the estimated sample average Fox News effect is increasing and more rich people voted for republicans but an the other hand the vote share for Republicans among rich people normally quite high anyways. Therefore it is hard to argue, that the implementation of Fox News is the main reason. Towns with a medium income or low income have decreasing sample average treatment effect.

```{r diffInDiffEducation, fig.cap="Differences in Vote Share in Average Higher and Lower Educated Towns", echo=FALSE, warning=FALSE}
=======

Figure \ref{fig:diffInDiffincome} focuses on how the average income in cities and the implementation of Fox News influence people's support for the Republican party. Those towns are considered to be high-income cities where the average income is over or equals 3.420$ and low income towns are where the average income is below $1.500. The results show that in high income towns the the support for the Republican party increased in both towns with Fox News and without even though the towns where Fox News was implemented already had higher Republican vote share. In low-income cities, the support for the Republican party increased in towns without or limited access to Fox News and slightly raised in towns where Fox News was implemented. Therefore, income does not seem to have a clear effect on whether the implementation of Fox News influenced the support for Republicans. 

```{r diffInDiffeducation, fig.cap="Differences in Vote Share in Average Higher and Lower Educated Towns", echo=FALSE, warning=FALSE}
>>>>>>> 2cc3803087b13bf2a496dd0f7d4db66cbc1ce6e8

high_edu.t_w_foxnews <- subset(towns_w_foxnews, subset = college1990 >= 0.2)
high_edu.t_no_foxnews<- subset(towns_no_foxnews, subset = college1990 >= 0.2)
high_edu.t_some_foxnews<- subset(towns_some_foxnews, subset = college1990 >= 0.2)

means_high_edu.t_w_foxnews<-colMeans(high_edu.t_w_foxnews[c('gopvoteshare1992','gopvoteshare1996','gopvoteshare2000')], na.rm = T)

means_high_edu.t_no_foxnews<-colMeans(high_edu.t_no_foxnews[c('gopvoteshare1992','gopvoteshare1996','gopvoteshare2000')], na.rm = T)

means_high_edu.t_some_foxnews<-colMeans(high_edu.t_some_foxnews[c('gopvoteshare1992','gopvoteshare1996','gopvoteshare2000')], na.rm = T)

low_edu.t_w_foxnews<- subset(towns_w_foxnews, subset = college1990 <= 0.08)
low_edu.t_no_foxnews<- subset(towns_no_foxnews, subset = college1990 <= 0.08)
low_edu.t_some_foxnews<- subset(towns_some_foxnews, subset = college1990 <= 0.08)

means_low_edu.t_w_foxnews<-colMeans(low_edu.t_w_foxnews[c('gopvoteshare1992','gopvoteshare1996','gopvoteshare2000')], na.rm = T)

means_low_edu.t_no_foxnews<-colMeans(low_edu.t_no_foxnews[c('gopvoteshare1992','gopvoteshare1996','gopvoteshare2000')], na.rm = T)

means_low_edu.t_some_foxnews<-colMeans(low_edu.t_some_foxnews[c('gopvoteshare1992','gopvoteshare1996','gopvoteshare2000')], na.rm = T)


par(mfrow = c(1, 2))
par(family = "serif")
plot(years, means_high_edu.t_w_foxnews, pch = 16, col = "darksalmon",
xlim = c(1992, 2000), ylim = c(0.2, 0.7), xlab = "Years of presidential election",
ylab = "Average Vote Share for Republicans",
main = "Higher Educated Population")


points(years, means_high_edu.t_no_foxnews,pch = 16, col = "deeppink4")
points(years, means_high_edu.t_some_foxnews,pch = 16, col = "grey80")


lines(years, means_high_edu.t_w_foxnews, type="c", col="darksalmon")
lines(years, means_high_edu.t_no_foxnews,type="c", col = "deeppink4")
lines(years, means_high_edu.t_some_foxnews,type="c", col = "grey80")


legend(x = 1992, y = 0.4, bty = "n",
       legend = c("Towns with Fox News", "Towns with some Fox News", "Towns without Fox News"),
       col = c("darksalmon", "deeppink4", "grey80"), pch = 19,
       pt.cex = c(0.5))

plot(years, means_low_edu.t_w_foxnews, pch = 16, col = "darksalmon",
xlim = c(1992, 2000), ylim = c(0.2, 0.7), xlab = "Years of presidential election",
ylab = "Average Vote Share for Republicans",
main = "Lower Educated Population")


points(years, means_low_edu.t_no_foxnews,pch = 16, col = "deeppink4")
points(years, means_low_edu.t_some_foxnews,pch = 16, col = "grey80")


lines(years, means_low_edu.t_w_foxnews, type="c", col="darksalmon")
lines(years, means_low_edu.t_no_foxnews,type="c", col = "deeppink4")
lines(years, means_low_edu.t_some_foxnews,type="c", col = "grey80")


```

<<<<<<< HEAD
The last measurement of finding the differences in sample average treatment effects by socioeconomic backgrounds is SATE for different levels of education. Here is it visible, that high educated regions are less likely to vote for republicans and have lower starting position. Looking at a medium or low level of education a decrease of the potential effect is visible, even though is a very small one.
=======

Finally, Figure \ref{fig:diffInDiffeducation} measures whether the average level of education in a city and access to Fox News influences people's support for the Republican party. A city is considered to be "higher educated" when 20% or more of the population has a college degree and it is considered to be "lower educated" when this number is smaller than 8%. The results show that in cities where the population tends to be lower educated, there is an increase in the support of the Republican party but in every city with or without Fox News as well. Interestingly, in towns where the population tends to be higher educated the increase of the Republican vote share is less intense.  
>>>>>>> 2cc3803087b13bf2a496dd0f7d4db66cbc1ce6e8

In summary, socioeconomic factors do not seem to have a clear effect on the influence of Fox News implementation on the increasing support for the Republican party.  The findings are not salient and decisive on the basis of various socio-economic criteria and thus no 'Fox News effect' was found. There have been changes that seem reasonable and appropriate, but it would be presumptuous to relate these to a Fox News effect, as the changes have always been marginal.

# Regression analysis

## Linear models

We already tested above the influence of different socioeconomic factors effects on people's support for the republican party and how the presence of Fox News in some cities changed people's party choices. However, we are also interested in to what extent socioeconomic factors or the potential effect of Fox News' presence in some cities influence people's vote for the Republican party. To answer this questions we will test our hypothesis (Fox news and the socioeconomic background affects the behavior of the electorate) through regression analysis.

```{r, results="asis", echo=FALSE, warning=FALSE}
#effect of social background

data$hisp_or_black <- data$black1990 + data$hisp1990

fox_news_effect <- lm(gopvoteshare2000 ~ subrf2000 + income1990 + hisp1990 + black1990 + college1990, data = data)


fox_news_effect_change <- lm(diff_1996_2000 ~ subrf2000 + income1990 + hisp1990 + black1990 + college1990, data = data)

# Combine the coefficient tables into a single table
combined_table <- rbind(cbind(summary(fox_news_effect)$coefficients[,1], summary(fox_news_effect_change)$coefficients[,1]), c(summary(fox_news_effect)$r.squared, summary(fox_news_effect_change)$r.squared), c(summary(fox_news_effect)$adj.r.squared, summary(fox_news_effect_change)$adj.r.squared), c(summary(fox_news_effect)$sigma, summary(fox_news_effect_change)$sigma), c("< 2.2e-16", "< 2.2e-16"))

# Rename the column names
colnames(combined_table) <- c("Effect of different factors \non Repubican vote share", "Effec of different factors \non the change of Republican vote share")

rownames(combined_table) <- c("Intercept", "Share of Fox-news \nsubscribers", "Level of education", "Share of black people", "Share of hispanic people", "Average income", "R^2", "Adjusted R^2", "Residual Std. Error", "P value")


library(knitr)
library(kableExtra)

# Print the combined table using kable
kable(combined_table, "latex")%>% 
  kable_styling(latex_options=c("scale_down","striped", stripe_index = c(1,2, 5:6)))

```

This is a linear regression model that aims to predict the vote share for republicans at the 2000 presidential election using four predictor variables: the share of Fox News subscribers in 2000, the median income in 1990, the proportion of hispanic or black population in 1990, and the proportion of population with a college degree in 1990. The model shows that the proportion of hispanic or black population and the proportion of population with a college degree are statistically significant predictors of the vote share for republicans, while the share of Fox News subscribers and median income are not significant predictors. The adjusted R-squared value of 0.1713 indicates that only about 17% of the variability in the vote share for republicans can be explained by the predictor variables included in the model.

## Prediction

When interpreting the results of a statistical analysis, it is important to consider not only the point estimates of the coefficients but also their associated confidence intervals. The confidence interval provides a range of plausible values for the true population parameter with a specified degree of confidence.

The 95% confidence interval for a coefficient estimate means that if we were to repeat the study many times, in 95% of the cases the true population parameter would fall within the given interval. Thus, it provides a measure of the precision and uncertainty of the estimate.

```{r foxFX, fig.show="hold", out.width="50%", fig.cap="Predicted effect of Fox News on Republican vote share and change in Republican vote share, including confidence interval.", echo=FALSE}
printPlot <- function(var1, var2, xlab, ylab, col){
  par(family = "serif")
  plot(c(var1[1], var2[1]) ~ c(0, 1), 
     ylim = c(min(c(var1[2], var2[2])), max(c(var1[3], var2[3]))), 
     type="l", 
     col=col, 
     xlab=xlab,
     ylab=ylab,
     lwd=3)
  polygon(c(0, 1, 1, 0), c(var1[2], var2[2], var2[3], var1[3]), col=adjustcolor(col, alpha.f = 0.2))
  
  abline(h=0, col="grey", lty=2)
}


fox_0 <- colMeans(predict (fox_news_effect, newdata = data.frame(subrf2000 = 0, income1990 = data$income1990, hisp1990 = data$hisp1990, black1990 = data$black1990, college1990 = data$college1990), interval = "confidence", level = 0.95))
fox_1 <- colMeans(predict (fox_news_effect, newdata = data.frame(subrf2000 = 1, income1990 = data$income1990, hisp1990 = data$hisp1990, black1990 = data$black1990, college1990 = data$college1990), interval = "confidence", level = 0.95))

printPlot(fox_0, fox_1, "Share of Fox News subscribers", "Predicted Republican vote share", "#bb9900")


printPlot <- function(var1, var2, xlab, ylab, col){
  par(family = "serif")
  plot(c(var1[1], var2[1]) ~ c(0, 1), 
     ylim = c(-0.08, 0.1), 
     type="l", 
     col=col, 
     xlab=xlab,
     ylab=ylab,
     lwd=3)
  polygon(c(0, 1, 1, 0), c(var1[2], var2[2], var2[3], var1[3]), col=adjustcolor(col, alpha.f = 0.2))
  abline(h=0, col="grey", lty=2)
}


change0 <- colMeans(predict (fox_news_effect_change, newdata = data.frame(subrf2000 = 0, income1990 = data$income1990, hisp1990 = data$hisp1990, black1990 = data$black1990, college1990 = data$college1990), interval = "confidence", level = 0.95))
change1 <- colMeans(predict (fox_news_effect_change, newdata = data.frame(subrf2000 = 1, income1990 = data$income1990, hisp1990 = data$hisp1990, black1990 = data$black1990, college1990 = data$college1990), interval = "confidence", level = 0.95))

printPlot(change0, change1, "Share of Fox News subscribers", "Predicted change in Republican vote share", "#22aa00")


```

Figure \ref{fig:foxFX} analyses the Foxnews effect both on the Republican vote share and on the change of Republican vote share between elections. The graphs show that as the share of fox news subscribers increase, both the predicted Republican vote share and the positive change of Republican vote share decreases, indicating a negative Foxnews effect. It is notable that at predicted Republican vote share the uncertainty range allows for no decrease or a minor increase, allowing no strong implications.

In order to contrast the measured Fox news effect with other possibly determining variables, the predicted effect of other socio-economic factors in electoral outcome in terms of change of Republican vote share between elections.

```{r incomeAndEducationFX, fig.show="hold", out.width="50%", fig.cap="Predicted effect of Income and Education on Republican vote share", echo=FALSE}

predIncome0 <- colMeans(predict (fox_news_effect_change, newdata = data.frame(subrf2000 = data$subrf2000, income1990 = 0, hisp1990 = data$hisp1990, black1990 = data$black1990, college1990 = data$college1990), interval = "confidence", level = 0.95), na.rm=T)
predIncome1 <- colMeans(predict (fox_news_effect_change, newdata = data.frame(subrf2000 = data$subrf2000, income1990 = 1, hisp1990 = data$hisp1990, black1990 = data$black1990, college1990 = data$college1990), interval = "confidence", level = 0.95), na.rm=T)

printPlot(predIncome0, predIncome1, "Level of Income", "Predicted Republican vote share", "#0000aa")

predEdu0 <- colMeans(predict (fox_news_effect_change, newdata = data.frame(subrf2000 = data$subrf2000, income1990 = data$income1990, hisp1990 = data$hisp1990, black1990 = data$black1990, college1990 = 0), interval = "confidence", level = 0.95), na.rm=T)
predEdu1 <- colMeans(predict (fox_news_effect_change, newdata = data.frame(subrf2000 = data$subrf2000, income1990 = data$income1990, hisp1990 = data$hisp1990, black1990 = data$black1990, college1990 = 1), interval = "confidence", level = 0.95), na.rm=T)

printPlot(predEdu0, predEdu1, "Level of Education", "Predicted Republican vote share", "#00ccaa")

```

```{r socioFX, fig.show="hold", out.width="50%", fig.cap="Predicted effect of  on Republican vote share", echo=FALSE}
predBlack0 <- colMeans(predict (fox_news_effect_change, newdata = data.frame(subrf2000 = data$subrf2000, income1990 = data$income1990, hisp1990 = data$hisp1990, black1990 = 0, college1990 = data$college1990), interval = "confidence", level = 0.95), na.rm=T)
predBlack1 <- colMeans(predict (fox_news_effect_change, newdata = data.frame(subrf2000 = data$subrf2000, income1990 = data$income1990, hisp1990 = data$hisp1990, black1990 = 1, college1990 = data$college1990), interval = "confidence", level = 0.95), na.rm=T)

printPlot(predBlack0, predBlack1, "Share of black population", "Predicted Republican vote share", "#882222")

predHisp0 <- colMeans(predict (fox_news_effect_change, newdata = data.frame(subrf2000 = data$subrf2000, income1990 = data$income1990, hisp1990 = 0, black1990 = data$black1990, college1990 = data$college1990), interval = "confidence", level = 0.95), na.rm=T)
predHisp1 <- colMeans(predict (fox_news_effect_change
                               , newdata = data.frame(subrf2000 = data$subrf2000, income1990 = data$income1990, hisp1990 = 1, black1990 = data$black1990, college1990 = data$college1990), interval = "confidence", level = 0.95), na.rm=T)

printPlot(predHisp0, predHisp1, "Share of hispanic population", "Predicted Republican vote share", "#9922ff")

```

Observing the different socio-economic factors clearly indicate certain tendencies, that are more far reaching than the previously seen fox-news effect. Figure \@ref(fig:incomeAndEducationFX) suggests that both lower income and lower educated neighborhoods have a higher likeliness to result in an increased Republican vote share, while both higher income and higher educated electoral districts are expected to have a slight decrease in Republican vote share. Both the intensity of the difference among the ends and the certainty of the prediciton is higher when observing the effect of the level of education.

Figure \@ref(fig:socioFX) indicates that areas with lower share of hispanic and black population are more likely to result in an increase in republican vote share while higher share of black and hispanic population are morel likely to have a slight decrease in Republican vote share. However, both the intensity and the uncertainty is higher of the effect of share of black population.

Consequently it is an important observation that the effect of the presence or absence of Fox-news is far less significant and far more uncertain than socio economic factors such as ethnic diversity or intellectual or material wealth. These findings highlight the importance of considering a broader range of variables when examining electoral outcomes and suggest that factors related to socio-economic conditions play a more substantial role in shaping voting patterns.

# Discussion & Conclusion

After the analysis, it can be summarized that 'a Fox News Effect' is not clearly visible and verifiable. Thus, the first hypothesis of the paper must also be refuted, as there is no clear evidence that Fox News together with socioeconomic background affects the behavior of the electorate in 2000. This was evident from the results of the regressions 'foxnews_2000' and 'foxnews_present_effect' which showed no significance or clear correlation. It was also evident from the sample average treatment affect studies that all of the effects shown can only be described as minimal.

The reasons why an effect is only so minimally visible can be manifold. First of all, it must be mentioned that a new channel may take a while to be added to a cable provider's lineup because of the talks negotiations during the implementation. Due of this, contrary to popular belief, Fox News was unable to establish itself as the first player in conservative media markets. this means that people who live in cities without fox news are still influenced by conservative TV channels, but are not recorded as such in this data set because they watch a different conservative channel. Furthermore, it is important to note that while there is a small effect, covering the external and internal variables that influence an election is extremely difficult. Third, the heavily conservative states of New Jersey and Wyoming would be the perfect example of this, ranking second and third, respectively, among the states with the largest vote share. This would mean that the effect could be weakened, since people have already voted conservatively anyway.

This would mean that the effect could be weakened, since people have already voted conservatively anyway. However, even if the aforementioned external problems were not present and the internal validity is flawless, it is possible that the Fox News effect would only play a minor role. As the theories and also the regression results have shown, socioeconomic factors play an important role. Thus, the second hypothesis of the paper can be confirmed, namely that socioeconomic factors are crucial for voting in the American two-party system. In particular, the factors of race and the level of education were the most decisive influences. Lastly, another potential explanation to the the general increase of support for the Republican party in most of the states is that the US had a democratic president (Bill Clinton) during this time period who had an impeachment. These external factors could have a more important positive effect on people's support for the Republican party than the influence of watching Fox News.

\newpage

# References
